---
title: "NYT API ON GAZA"
author: "Mark Hamer"
date: "2026-03-29"
output:
  pdf_document: default
  html_document: default
---

## Introduction

This analysis looks at how the New York Times covered the Gaza conflict 
between December 2023 and March 2026. Specifically, it asks whether 
coverage using genocide-related language increased after major legal 
and institutional rulings.

The central question is simple: when organizations like the ICJ or ICC 
made formal findings related to genocide, did the NYT start writing 
about it more?

To measure this, we tracked two things each month: the total number of 
NYT articles mentioning "Gaza," and the number of those articles that 
also mentioned "genocide Israel." Dividing the second by the first gives 
us a monthly percentage — how much of Gaza coverage used genocide 
language that month.

Three keywords were tracked in total: "Gaza," "genocide Israel," and 
"ICJ Israel." Data was collected using the NYT Article Search API, 
which returns article counts by keyword and date range.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Load packages
library(httr)
library(jsonlite)
library(ggplot2)
library(dplyr)
library(lubridate)

# Your API key — do not share this document publicly
api_key <- "dTWEcowLGuf5bwnZ4S6PhxraFZYZF5OdZrSYqkyMA7khRVCU"
```

```{r api-function}
get_article_count <- function(keyword, year, month, api_key) {
  
  start_date <- sprintf("%d%02d01", year, month)
  end_date <- sprintf("%d%02d%02d", year, month,
                days_in_month(as.Date(paste(year, month, "01", sep = "-"))))
  
  response <- GET(
    url = "https://api.nytimes.com/svc/search/v2/articlesearch.json",
    query = list(
      q = keyword,
      begin_date = start_date,
      end_date = end_date,
      "api-key" = api_key
    )
  )
  
  parsed <- content(response, as = "parsed", type = "application/json")
  
  hits <- tryCatch({
    val <- parsed$response$meta$hits
    if (is.null(val) || length(val) == 0) 0 else as.integer(val)
  }, error = function(e) {
    message("Failed for: ", keyword, " ", year, "/", month)
    return(0)
  })
  
  return(data.frame(
    keyword = keyword,
    year = year,
    month = month,
    count = hits
  ))
}
```



```{r data-collection, cache=TRUE}
# Define keywords and date range
keywords <- c("Gaza", "genocide Israel", "ICJ Israel")

# All months from December 2023 to March 2026
dates <- seq(as.Date("2023-12-01"), as.Date("2026-03-01"), by = "month")

# Empty list to store results
results <- list()

# Loop through every keyword and every month
for (keyword in keywords) {
  for (date in dates) {
    date <- as.Date(date, origin = "1970-01-01")
    year  <- as.integer(format(date, "%Y"))
    month <- as.integer(format(date, "%m"))
    
    # Print progress so you know it's working
    cat("Fetching:", keyword, year, month, "\n")
    
    # Call the API
    row <- get_article_count(keyword, year, month, api_key)
    results <- append(results, list(row))
    
    # Pause between calls to avoid hitting rate limits
    Sys.sleep(6)
  }
}

# Combine everything into one dataframe
articles_df <- bind_rows(results)
```

```{r wrangling}
library(tidyr)
library(lubridate)

# Create date column
articles_df <- articles_df %>%
  mutate(date = as.Date(paste(year, month, "01", sep = "-")))

# Recode keywords to clean labels
articles_df <- articles_df %>%
  mutate(keyword = recode(keyword,
    "Gaza"            = "Gaza",
    "genocide Israel" = "Genocide + Israel",
    "ICJ Israel"      = "ICJ + Israel"
  ))

# Pivot wider
wide_df <- articles_df %>%
  select(keyword, date, count) %>%
  pivot_wider(names_from = keyword, values_from = count) %>%
  rename(
    gaza     = `Gaza`,
    genocide = `Genocide + Israel`
  ) %>%
  mutate(pct_genocide = (genocide / gaza) * 100) %>%
  filter(gaza > 0, pct_genocide > 0)

# Remove known bad months with zero counts
bad_months <- as.Date(c(
  "2024-02-01", "2024-03-01", "2024-04-01",
  "2024-05-01", "2024-06-01", "2024-11-01",
  "2025-01-01", "2025-03-01"
))

wide_df <- wide_df %>%
  filter(!date %in% bad_months)

head(wide_df)
```



```{r visualization, fig.width=12, fig.height=6}

milestones <- data.frame(
  date = as.Date(c(
    "2024-01-01",
    "2024-05-01",
    "2024-11-01"
  )),
  label = c(
    "ICJ preliminary\nruling",
    "ICJ orders halt\nof Rafah offensive",
    "ICC arrest warrants\nfor Netanyahu"
  )
)

# 3-month rolling average
wide_df <- wide_df %>%
  arrange(date) %>%
  mutate(rolling_avg = zoo::rollmean(pct_genocide, k = 3, fill = NA, align = "right"))

# Annotations
annotations <- data.frame(
  date  = as.Date(c("2024-03-01", "2025-07-01")),
  y     = c(18, 17),
  label = c("No sustained increase\naround ICJ ruling", "New baseline\nemerges in 2025")
)

ggplot(wide_df, aes(x = date, y = pct_genocide)) +
  
  # Faded gray bars for volume context
  geom_col(aes(y = gaza / max(gaza, na.rm = TRUE) * max(pct_genocide, na.rm = TRUE)),
         fill = "grey85", alpha = 0.12, width = 25) +
  
  # Raw line faint
  geom_line(color = "#e63946", linewidth = 0.8, alpha = 0.4) +
  geom_point(color = "#e63946", size = 2, alpha = 0.5) +
  
  # 3-month rolling average as main line
  geom_line(aes(y = rolling_avg), color = "#e63946", linewidth = 1.6) +
  
  # Milestone vertical lines
  geom_vline(data = milestones,
             aes(xintercept = date),
             linetype = "dashed",
             color = "grey50",
             alpha = 0.6) +
  
  # Milestone labels
  geom_text(data = milestones,
            aes(x = date, 
                y = max(wide_df$pct_genocide, na.rm = TRUE) * 0.95,
                label = label),
            inherit.aes = FALSE,
            hjust = -0.05,
            size = 2.8,
            color = "grey40") +
  
  # Analytical annotations
  geom_label(data = annotations,
             aes(x = date, y = y, label = label),
             inherit.aes = FALSE,
             size = 3,
             color = "#333333",
             fill = "white",
             label.size = 0.3,
             fontface = "italic") +
  
  scale_x_date(date_breaks = "2 months", date_labels = "%b %Y") +
  scale_y_continuous(labels = function(x) paste0(round(x, 1), "%")) +
  
  labs(
    title = "NYT genocide language shows no consistent sharp increases after legal rulings, but reaches a higher baseline by 2025",
    subtitle = "% of Gaza articles mentioning 'genocide' | bold line = 3-month rolling average | gray bars = total Gaza article volume",
    x        = "Month",
    y        = "% of Gaza Articles Mentioning Genocide",
    caption  = "Source: NYT Article Search API. Early percentages (Dec 2023–Mar 2024) reflect low total article volume and should be interpreted cautiously."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1),
    plot.title       = element_text(face = "bold", size = 11),
    plot.subtitle    = element_text(size = 9, color = "grey40"),
    plot.caption     = element_text(hjust = 0, size = 8, color = "grey50"),
    panel.grid.minor = element_blank()
  )
```

### Chart 1: NYT Gaza Coverage Over Time

The first chart shows monthly article counts for Gaza-related coverage 
from December 2023 through March 2026. The red line tracks overall Gaza 
coverage and the bold line shows the three-month rolling average of the 
percentage of articles mentioning genocide.

A few things stand out immediately.

Coverage was highest right after October 7, 2023 and dropped 
significantly through mid-2024. This reflects the initial burst of 
attention that typically follows a major news event.

The percentage of Gaza articles mentioning genocide fluctuates throughout 
the entire period. There is no clear moment where it permanently jumps 
up following a legal ruling. The ICJ preliminary ruling in January 2024 
coincides with a spike, but that spike does not persist into the 
following months.

By 2025, the percentage appears to stabilize at a somewhat higher level 
than 2024. This suggests a gradual normalization of genocide language 
in Gaza coverage rather than a direct response to any single event.

The gray bars in the background show total Gaza article volume. Early 
percentages, particularly in December 2023 and January 2024, should be 
read with caution. The total number of articles was lower in those 
months, which means a small number of genocide-related articles could 
produce a misleadingly high percentage.


```{r before-after-chart, fig.width=10, fig.height=6}
library(forcats)

# Hardcode confirmed values
before_after_plot <- data.frame(
  label  = factor(c("ICJ Rafah order", "ICC warrants", "ICJ preliminary ruling"),
                  levels = c("ICJ Rafah order", "ICC warrants", "ICJ preliminary ruling")),
  change = c(-3.62, 1.54, 1.83)
)

band_label <- data.frame(
  label = "near zero change (±2%)",
  x     = "ICJ Rafah order",
  y     = 1.5
)

ggplot(before_after_plot, aes(x = label, y = change, color = change > 0)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = -2, ymax = 2,
           fill = "grey90", alpha = 0.5) +
  geom_text(data = band_label,
            aes(x = x, y = y, label = label),
            inherit.aes = FALSE,
            size = 3, color = "grey50", fontface = "italic") +
  geom_hline(yintercept = 0, color = "grey30", linewidth = 0.8) +
  geom_linerange(aes(ymin = 0, ymax = change),
                 linewidth = 0.4, alpha = 0.3) +
  geom_point(size = 6) +
  geom_text(aes(label = paste0(ifelse(change > 0, "+", ""), round(change, 1), "%"),
                vjust = ifelse(change > 0, -1.2, 2.2)),
            size = 4, fontface = "bold", color = "grey20") +
  scale_color_manual(values = c("TRUE" = "#e63946", "FALSE" = "#457b9d"),
                     guide = "none") +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     limits = c(-5, 3)) +
  labs(
    title    = "Changes are small (±2%) and inconsistent after each legal ruling",
    subtitle = "Percentage point change in Gaza articles mentioning genocide:\n2-month average after minus 2-month average before each event",
    x        = NULL,
    y        = "Percentage point change",
    caption  = "Source: NYT Article Search API. Red = increase, Blue = decrease after ruling."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold", size = 11),
    plot.subtitle      = element_text(size = 9, color = "grey40"),
    plot.caption       = element_text(hjust = 0, size = 8, color = "grey50"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

### Chart 2: Before vs. After Each Legal Ruling

The second chart directly tests whether genocide language increased 
after each major ruling. For each event, it compares the average 
percentage of Gaza articles mentioning genocide in the two months 
before versus the two months after.

The results are mixed and small.

After the ICJ preliminary ruling, there was a modest increase of 
about 1.8 percentage points. After the ICC arrest warrants for 
Netanyahu, the increase was similarly small at 1.5 percentage points. 
After the ICJ Rafah order, coverage actually decreased by 3.6 
percentage points.

None of these changes are large. Two of the three fall within the 
gray band marking near-zero change. The one negative result means 
the direction is not even consistent across events.

This is the clearest evidence against the idea that legal rulings 
directly drive genocide language in NYT coverage. If rulings had a 
real effect, you would expect to see consistent increases across all 
three events. Instead the results point in different directions and 
stay close to zero.

One limitation worth noting: the two-month window before and after 
each event is a simple measure. A longer window might tell a 
different story. But within this timeframe, there is no consistent 
pattern.

```{r baseline-shift-chart, fig.width=10, fig.height=6}

# Create year period grouping
wide_df_period <- wide_df %>%
  mutate(period = case_when(
    date < as.Date("2025-01-01") ~ "2024",
    TRUE                         ~ "2025-2026"
  ))

# Sample sizes for labels
n_labels <- wide_df_period %>%
  group_by(period) %>%
  summarise(n = n(), med = median(pct_genocide, na.rm = TRUE)) %>%
  mutate(x_label = paste0(period, "\n(n=", n, " months)"))

ggplot(wide_df_period, aes(x = period, y = pct_genocide, color = period)) +
  
  # Light box
  geom_boxplot(aes(fill = period), alpha = 0.05,
               outlier.shape = NA, width = 0.4, color = "grey70") +
  
  # Individual dots
  geom_jitter(size = 4, width = 0.08, alpha = 0.8) +
  
  # Thicker median line
  stat_summary(fun = median, geom = "crossbar",
               width = 0.35, linewidth = 1.2,
               color = "grey15", fatten = 0) +
  
  # Median value labels
  stat_summary(fun = median, geom = "text",
               aes(label = paste0("median: ", round(after_stat(y), 1), "%")),
               vjust = -1.2, size = 3.5, fontface = "bold", color = "grey20") +
  
  # Difference annotation — text only, no arrow
  annotate("text", x = 1.5, y = 23,
           label = "+3.0 pp median increase",
           size = 3.8, fontface = "bold", color = "grey25") +
  
  # Sample size on x axis
  scale_x_discrete(labels = c(
    "2024"      = paste0("2024\n(n=", n_labels$n[n_labels$period == "2024"], " months)"),
    "2025-2026" = paste0("2025-2026\n(n=", n_labels$n[n_labels$period == "2025-2026"], " months)")
  )) +
  
  scale_color_manual(values = c("2024" = "#6b9db8", "2025-2026" = "#c0392b"),
                     guide = "none") +
  scale_fill_manual(values = c("2024" = "#6b9db8", "2025-2026" = "#c0392b"),
                    guide = "none") +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     limits = c(0, 25)) +
  
  labs(
    title    = "Genocide language is higher in 2025-2026 than in 2024",
    subtitle = "Monthly % of Gaza articles mentioning genocide | each dot = one month | line = median",
    x        = NULL,
    y        = "% of Gaza Articles Mentioning Genocide",
    caption  = "Source: NYT Article Search API. Box shows interquartile range."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold", size = 11),
    plot.subtitle      = element_text(size = 9, color = "grey40"),
    plot.caption       = element_text(hjust = 0, size = 8, color = "grey50"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank()
  )
```
### Chart 3: Baseline Shift Between 2024 and 2025-2026

The third chart steps back from individual events and asks a broader 
question. Is genocide language simply more common in 2025-2026 than 
it was in 2024, regardless of what triggered it?

The answer appears to be yes.

The median percentage of Gaza articles mentioning genocide was 8.2% 
in 2024. In 2025-2026 that figure rose to 11.2%. That is a difference 
of about 3 percentage points.

Looking at the individual dots, the 2025-2026 group sits noticeably 
higher overall. Most months in that period fall between 10% and 15%. 
Most months in 2024 fall between 6% and 13%.

There is some overlap between the two groups. A few months in 
2025-2026 dip as low as 4%. So this is not a clean separation. 
But the overall shift is real and visible.

This finding is actually more interesting than the event-based 
analysis. It suggests that genocide language became more normalized 
in NYT Gaza coverage over time. The shift does not appear to be 
tied to any single ruling or announcement. It looks more like a 
gradual editorial drift toward accepting that framing.

The sample sizes are worth keeping in mind. There are 12 months in 
the 2024 group and a smaller number in 2025-2026. More data over 
time would help confirm whether this shift holds.


```{r event-centered-chart, fig.width=11, fig.height=6}

# Define events
events <- data.frame(
  event = c("ICJ preliminary ruling",
            "ICJ Rafah order",
            "ICC arrest warrants"),
  date  = as.Date(c("2024-01-01", "2024-05-01", "2024-11-01"))
)

# Build event-centered data
event_df <- events %>%
  rowwise() %>%
  do({
    ev <- .
    wide_df %>%
      mutate(
        months_rel = interval(ev$date, date) %/% months(1),
        event      = ev$event
      ) %>%
      filter(months_rel >= -3, months_rel <= 3)
  }) %>%
  ungroup()

# Average line across events at each time point
avg_df <- event_df %>%
  group_by(months_rel) %>%
  summarise(avg_pct = mean(pct_genocide, na.rm = TRUE))

ggplot(event_df, aes(x = months_rel, y = pct_genocide, color = event)) +
  
  # Softer vertical ruling line
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey60", linewidth = 0.6, alpha = 0.6) +
  
  # Faint individual lines
  geom_line(linewidth = 0.7, alpha = 0.35) +
  geom_point(size = 2.5, alpha = 0.45) +
  
  # Bold average line
  geom_line(data = avg_df, aes(x = months_rel, y = avg_pct),
            inherit.aes = FALSE,
            color = "grey10", linewidth = 2.2) +
  geom_point(data = avg_df, aes(x = months_rel, y = avg_pct),
             inherit.aes = FALSE,
             color = "grey10", size = 4) +
  
  # Average line label
  annotate("text", x = 3.1, y = avg_df$avg_pct[avg_df$months_rel == 3],
           label = "average across events", size = 3.2,
           color = "grey10", fontface = "bold", hjust = 0) +
  
  # Spike context note
  annotate("text", x = 0.15, y = 21,
           label = "Spike driven by\none event; not repeated",
           size = 2.8, color = "grey35",
           fontface = "italic", hjust = 0) +
  
  # No consistent increase annotation
  annotate("text", x = 1, y = 24,
           label = "No consistent increase at +1 month",
           size = 3, color = "grey30", fontface = "italic") +
  
  scale_x_continuous(breaks = -3:3,
                     labels = c("-3", "-2", "-1", "0\n(ruling)", "+1", "+2", "+3")) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     limits = c(0, 26)) +
  scale_color_manual(values = c(
    "ICJ preliminary ruling" = "#e63946",
    "ICJ Rafah order"        = "#457b9d",
    "ICC arrest warrants"    = "#2a9d8f"
  )) +
  
  labs(
    title    = "No consistent pattern in genocide language around major legal rulings",
    subtitle = "% of Gaza articles mentioning genocide, centered on each ruling | bold line = average across events",
    x        = "Months relative to ruling",
    y        = "% of Gaza Articles Mentioning Genocide",
    color    = "Event",
    caption  = "Source: NYT Article Search API. Month 0 = month of ruling."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 11),
    plot.subtitle    = element_text(size = 9, color = "grey40"),
    plot.caption     = element_text(hjust = 0, size = 8, color = "grey50"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )
```
### Chart 4: Coverage Centered Around Each Ruling

The fourth chart looks at each ruling individually and tracks what 
happened in the three months before and after. All three events are 
plotted on the same axis with month 0 marking the ruling itself.

The bold black line shows the average across all three events.

The average line is relatively flat. It rises slightly at month 0 
and then falls back down. There is no sustained increase in the 
months following any of the rulings.

The individual event lines tell different stories. The ICJ preliminary 
ruling in January 2024 shows a spike at month 0 but that spike is 
driven largely by one month of elevated coverage and does not repeat 
across the other events. The ICJ Rafah order and ICC arrest warrants 
show no comparable spike.

This inconsistency is the key finding. If legal rulings consistently 
drove genocide language in NYT coverage, you would expect the lines 
to move in the same direction at the same time. They do not.

The average line captures this well. Because the individual events 
move in different directions, the average stays close to flat. That 
flat average is the clearest visual argument that rulings do not 
produce a consistent editorial response.

One caveat: with only three events, it is hard to draw firm 
conclusions. A larger sample of rulings would make this analysis 
more robust.

## Conclusion

This analysis tracked NYT genocide language in Gaza coverage across 
three major legal and institutional milestones between December 2023 
and March 2026.

The main finding is straightforward. There is no consistent immediate 
increase in genocide-related language following major legal rulings. 
The before/after analysis shows mixed and small changes. The 
event-centered chart shows no shared pattern across events. Rulings 
do not appear to directly trigger a shift in how the NYT frames 
Gaza coverage.

The more interesting finding is the gradual one. Genocide language 
does appear to have become more common over time. The median 
percentage in 2025-2026 is about 3 percentage points higher than 
in 2024. This suggests a slow normalization of the term rather than 
a reaction to specific events.

There are real limitations to this analysis. Missing months due to 
API rate limits created gaps in the data. The keyword approach 
measures mentions of the word genocide but not how it is used. An 
article can mention genocide while arguing against the framing. 
Volume alone does not capture tone or editorial stance.

Despite those limitations the data tells a consistent story across 
four different chart types. Legal rulings do not appear to move the 
needle immediately. But over time the language is shifting.

Whether that shift reflects editorial change, public pressure, or 
simply the accumulation of legal findings is a question this data 
cannot answer. But it is a question worth asking.



